---
title: "snRNA_co-expression_genes"
author: "Kristina Benevides"
date: "2025-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r warning=FALSE, message=FALSE}
library(here)
library(readr)
library(scLink)
library(pheatmap)
library(ggplot2)
library(SingleCellExperiment)
library(Matrix)
library(tidyr)
library(Seurat)
```

## Load the data
```{r load_data}
sce <- readRDS(here("data/sce_neuron.rds"))
neuron_seurat <- readRDS(here("data/neuron_seurat_2.rds"))
expr_matrix_all <- readRDS(here("data/expr_matrix_all.rds"))

neuron_types = c("chat1","chat2","chat3","Glu1","Glu2","Glu3","Glu4","Glu5","Glu6",
                 "Glu7","Glu8","Glu9","Glu10","Glu11","Glu12","Glu13","Glu14","Glu15",
                 "GABA1","GABA2","GABA3","GABA4","GABA5","GABA6","GABA7")
```

## If you first want to quickly check overall expression for two genes
```{r message=FALSE, warning=FALSE}

genes_subset <- c("Gfral","Tlr3") # CHANGE

ft1 <- FeaturePlot(neuron_seurat, features = genes_subset, min.cutoff = "q10")
ft2 <- FeaturePlot(neuron_seurat, features = genes_subset, blend = TRUE, pt.size = 0.1) +
  theme(plot.title = element_text(hjust = 0.5))

print(ft1)
print(ft2)

ggsave(here("img/expression_GOIs_neurons.png"), plot = ft1, width = 10, height = 6, dpi = 300)
ggsave(here("img/Co-expression_GOIs_neurons.png"), plot = ft2, width = 20, height = 12, dpi = 600)
```

## Genes of Interest

Set your genes of interest.

```{r genes, echo=FALSE}

genes <- c("Gfral","Tlr3", "Adcyap1", "Dbh") # CHANGE

# If you have more than a few genes, then put them in a file "GOIs.txt" (one gene per row) and use the below code
#genes <- readLines(here("GOIs.txt"))
```

## Co-expression GOIs

Run scLink to look at the co-expression of the GOIs. OBS! This step is very resource demanding, make sure to have at least 80 GB RAM on your system. 

```{r scLink, echo=FALSE}

count <- t(expr_matrix_all)

count.norm <- sclink_norm(count, scale.factor = 1e6, filter.genes = FALSE, gene.names = genes)
gc()

networks <- sclink_net(expr = count.norm, ncores = 1, lda = seq(0.5, 0.1, -0.05))

```

# Results from from scLink
```{r}
# Network correlation
networks$cor[1:length(genes),1:length(genes)]
# Network details
net1 = networks$summary[[1]]
net1
```

# Pairwise Co-expression values per neuron cell type
```{r}
# Run scLink per cell type

process_neuron_cell_type <- function(neuron_seurat, cell_type, genes) {
  # Ensure the active identity class is set to the clustering result
  Idents(neuron_seurat) <- "neuron_cell_type"  
  
  # Subset the Seurat object to include the specified neuron cell type
  neuron_seurat_subset <- subset(neuron_seurat, idents = cell_type)
  
  # Extract data layers
  data_layers <- grep("^data\\.", Layers(neuron_seurat_subset[["RNA"]]), value = TRUE)
  
  # Extract expression data from all layers and combine
  expr_list <- lapply(data_layers, function(layer) {
    layer_data <- neuron_seurat_subset[["RNA"]]@layers[[layer]]
    
    # Check if layer data is of type S4 and not NULL
    if (!is.null(layer_data)) {
      if (inherits(layer_data, "matrix") || inherits(layer_data, "dgCMatrix")) {
        mat <- GetAssayData(neuron_seurat_subset, assay = "RNA", layer = layer)
        
        cat("Processing layer:", layer, "with dimensions:", dim(mat), "\n")
        
        # Check if the matrix has two dimensions (genes x cells)
        if (length(dim(mat)) == 2) {
          return(mat)
        } else {
          warning(paste("Layer", layer, "does not have the expected two dimensions."))
          return(NULL)  # Skip this layer if it doesn't meet the criteria
        }
      } else {
        warning(paste("Layer", layer, "is not of type S4 or is unexpected. Skipping."))
        return(NULL)  # Skip this layer if it's not an S4 object
      }
    } else {
      warning(paste("Layer", layer, "is NULL. Skipping."))
      return(NULL)  # Skip this layer if it's NULL
    }
  })
  
  # Remove NULL entries from the list
  expr_list <- Filter(Negate(is.null), expr_list)
  
  # Check if there are any valid expression matrices
  if (length(expr_list) == 0) {
    stop(paste("No valid expression matrices found for cell type:", cell_type))
  }
  
  # Merge all extracted expression matrices (column-wise: cells are in columns)
  expr_matrix <- do.call(cbind, expr_list)
  
  # Transpose
  count <- t(expr_matrix)
  
  # Normalize counts
  count.norm <- sclink_norm(count, scale.factor = 1e6, filter.genes = FALSE, gene.names = genes)
  
  gc()
  
  # Create networks
  networks <- tryCatch({
    sclink_net(expr = count.norm, ncores = 1, lda = seq(0.5, 0.1, -0.05))
  }, error = function(e) {
    warning(paste("Error in creating networks for cell type:", cell_type, "\n", e$message))
    return(NULL)
  })
  
  if (!is.null(networks)) {
    return(networks$cor[1:length(genes), 1:length(genes)])
  } else {
    return(NULL)  # Return NULL if network creation failed
  }
}

# Loop over all neuron cell types
results <- lapply(neuron_types, function(cell_type) {
  cat("Processing cell type:", cell_type, "\n")
  process_neuron_cell_type(neuron_seurat, cell_type, genes)
})

# Optionally, name the results for easy identification
names(results) <- neuron_types

```
# Select subset of the above genes for pairwise comparisons
```{r}
gene1 <- "Gfral" # CHANGE
gene2 <- "Tlr3" # CHANGE  
```

# Pairwise Co-expression values per cell type
```{r}
# Initialize a data frame to hold co-expression values
co_expression_values <- data.frame(CellType = character(), CoExpression = numeric(), stringsAsFactors = FALSE)

# Loop through each entry in results
for (cell_type in names(results)) {
  # Extract the correlation matrix for the current cell type
  cor_matrix <- results[[cell_type]]
  
  # Check if the matrix contains both genes
  if (gene1 %in% rownames(cor_matrix) && gene2 %in% colnames(cor_matrix)) {
    # Extract the co-expression value for the GOIs
    co_expression_value <- cor_matrix[gene1, gene2]
    
    # Append the result to the data frame
    co_expression_values <- rbind(co_expression_values, data.frame(CellType = cell_type, CoExpression = co_expression_value))
  }
}
co_expression_values
```


# Scatter plot
```{r}

# Define clusters of interest
clusters_of_interest <- c("Glu8","Glu5", "Glu4", "Glu1") # CHANGE

# Define genes of interest
genes_of_interest <- genes

# Define subset
subset <- c(gene1, gene2)

# Extract logcounts for only selected genes
logcounts_data <- as.data.frame(t(as.matrix(logcounts(sce)[genes_of_interest, ])))

# Get metadata with cluster labels
metadata <- colData(sce)[, "predicted_cell_type", drop = FALSE]

# Combine expression data with metadata
subset_data <- cbind(logcounts_data, metadata)

# Filter for the clusters of interest
subset_data <- subset_data[subset_data$predicted_cell_type %in% clusters_of_interest, ]

# Convert to long format for ggplot
subset_data_long <- pivot_longer(subset_data, cols = subset, names_to = "Gene", values_to = "Expression")

# Create scatter plot
g <- ggplot(subset_data, aes(x = Gfral, y = Tlr3, color = predicted_cell_type)) + # CHANGE
  geom_point(alpha = 0.5) +
  facet_wrap(~predicted_cell_type) +
  theme_minimal() +
  labs(title = "Gfral vs. Tlr3 Expression",  #CHANGE
       x = "Gfral Expression (logcounts)",  #CHANGE
       y = "Tlr3 Expression (logcounts)") #CHANGE
g

ggsave(here("img/GOIs_scatter_plots.png"), g, bg = 'white')
```

# Plot co-expression values for gene1/gene2 per cell type 
```{r message=FALSE, warning=FALSE}
# Extract gene2 correlation values from the gene1 row
genes_values <- sapply(results, function(x) {
  val <- x[gene1, gene2]
  if (is.null(val)) NA else as.numeric(val)  # Convert NULLs to NA and ensure numeric
})

# Convert to a 1-row numeric matrix
genes_matrix <- matrix(genes_values, nrow = 1)
rownames(genes_matrix) <- "Gfral-Tl3r" #CHANGE
colnames(genes_matrix) <- names(results)

# Replace NA values with 0 or use na.color argument in pheatmap
genes_matrix[is.na(genes_matrix)] <- 0  

# Create heatmap
p <- pheatmap(
  genes_matrix,
  cluster_rows = FALSE,  # No need to cluster one row
  cluster_cols = TRUE,   # Cluster cell types
  color = colorRampPalette(c("blue", "white", "red"))(100),
  main = "Gfral-Tlr3 Co-expression Across Cell Types", #CHANGE
  na_col = "gray", 
  angle_col = 45,
  cellheight = 60
  
)

ggsave(here("img/heatmap_cluster_coexp_tlr3.png"), p)
```

# Print co-expression values for only one cluster
```{r message=FALSE, warning=FALSE}
cluster <- "Glu8" #CHANGE
p_cluster <- pheatmap(results[[cluster]])
print(p_cluster)
ggsave(here("img/heatmap_cluster.png"),p_cluster)

```
# Plot co-expression values for gene1/gene2 per cell type 
```{r message=FALSE, warning=FALSE}
# Define your gene pairs
gene_pairs <- list(
  c("Gfral", "Tlr3"),
  c("Gfral", "Adcyap1"),
  c("Gfral", "Dbh")
)

# Optional: give rownames for gene pairs (for the heatmap rows)
pair_names <- sapply(gene_pairs, function(p) paste(p[1], p[2], sep = "-"))

# Create matrix: rows = gene pairs, cols = cell types (length of results)
genes_matrix <- do.call(rbind, lapply(gene_pairs, function(pair) {
  vals <- sapply(results, function(x) {
    val <- x[pair[1], pair[2]]
    if (is.null(val)) NA else as.numeric(val)
  })
  vals
}))

# Set row and column names
rownames(genes_matrix) <- pair_names
colnames(genes_matrix) <- names(results)

# Handle NA values (e.g., replace with 0)
genes_matrix[is.na(genes_matrix)] <- 0

# Symmetric color scale limits
max_val <- max(abs(genes_matrix), na.rm = TRUE)
breaks <- seq(-max_val, max_val, length.out = 100)

# Calculate plot dimensions dynamically
n_rows <- nrow(genes_matrix)
n_cols <- ncol(genes_matrix)
plot_width <- 1 + 0.3 * n_cols + 2  # Added more width (2 instead of 1.5) for color bar
plot_height <- 1 + 0.5 * n_rows

# Save heatmap with enough width to fit color bar
pheatmap(
  genes_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1),
  breaks = breaks,
  main = "Gene Pair Co-expression Across Cell Types",
  na_col = "gray",
  angle_col = 45,
  cellheight = 20,
  cellwidth = 20,
  fontsize = 10,
  filename = here("img/heatmap_cluster_coexp_multiple_genes_fullrange.png"),
  width = plot_width,
  height = plot_height
)

# Cap values to improve color contrast
cap_value <- 0.1  # You can adjust this depending on your data spread
genes_matrix_capped <- genes_matrix
genes_matrix_capped[genes_matrix_capped > cap_value] <- cap_value
genes_matrix_capped[genes_matrix_capped < -cap_value] <- -cap_value

# Define color scale
breaks <- seq(-cap_value, cap_value, length.out = 100)

# Dynamically calculate plot size
# Adjust these factors as needed for clarity
n_rows <- nrow(genes_matrix_capped)
n_cols <- ncol(genes_matrix_capped)
plot_width <- 1 + 0.3 * n_cols + 1.5  # +1.5 for space for color bar
plot_height <- 1 + 0.5 * n_rows

# Create heatmap
p <- pheatmap(
  genes_matrix_capped,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1),
  breaks = breaks,
  main = "Gene Pair Co-expression Across Cell Types",
  na_col = "gray",
  angle_col = 45,
  cellheight = 20,
  cellwidth = 20,
  fontsize = 10,
  filename = here("img/heatmap_cluster_coexp_multiple_genes.png"),
  width = plot_width,
  height = plot_height
)


```

# Cap values to improve color contrast
cap_value <- 0.1  # You can adjust this depending on your data spread
genes_matrix_capped <- genes_matrix
genes_matrix_capped[genes_matrix_capped > cap_value] <- cap_value
genes_matrix_capped[genes_matrix_capped < -cap_value] <- -cap_value

# Define color scale
breaks <- seq(-cap_value, cap_value, length.out = 100)

# Dynamically calculate plot size
# Adjust these factors as needed for clarity
n_rows <- nrow(genes_matrix_capped)
n_cols <- ncol(genes_matrix_capped)
plot_width <- 1 + 0.3 * n_cols + 1.5  # +1.5 for space for color bar
plot_height <- 1 + 0.5 * n_rows

# Create heatmap
p <- pheatmap(
  genes_matrix_capped,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1),
  breaks = breaks,
  main = "Gene Pair Co-expression Across Cell Types",
  na_col = "gray",
  angle_col = 45,
  cellheight = 20,
  cellwidth = 20,
  fontsize = 10,
  filename = here("img/heatmap_cluster_coexp_multiple_genes.png"),
  width = plot_width,
  height = plot_height
)
