---
title: "snRNA_co-expression_genes"
author: "Kristina Benevides"
date: "2025-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r warning=FALSE, message=FALSE}
library(here)
library(readr)
library(scLink)
library(pheatmap)
library(ggplot2)
library(SingleCellExperiment)
library(Matrix)
library(tidyr)
library(Seurat)
```

## Load the data
```{r load_data}
sce <- readRDS(here("data/sce_neuron.rds"))
neuron_seurat <- readRDS(here("data/neuron_seurat_2.rds"))
expr_matrix_all <- readRDS(here("data/expr_matrix_all.rds"))

# Make sure we’re on the RNA assay
DefaultAssay(neuron_seurat) <- "RNA"

# IMPORTANT for v5: merge multiple layers into one
neuron_seurat <- JoinLayers(neuron_seurat)

neuron_types = c("chat1","chat2","chat3","Glu1","Glu2","Glu3","Glu4","Glu5","Glu6",
                 "Glu7","Glu8","Glu9","Glu10","Glu11","Glu12","Glu13","Glu14","Glu15",
                 "GABA1","GABA2","GABA3","GABA4","GABA5","GABA6","GABA7")
```

## If you first want to quickly check overall expression for two genes
```{r message=FALSE, warning=FALSE, dev = "svg"}

genes_subset <- c("Gfral","Tmtc2", "Slc6a15", "Man1a", "Akap13", "Etv1", "Tlr3") # CHANGE

ft1 <- FeaturePlot(neuron_seurat, features = genes_subset, min.cutoff = "q10")
ft2 <- FeaturePlot(neuron_seurat, features = c("Gfral", "Tlr3"), blend = TRUE, pt.size = 0.1) +
  theme(plot.title = element_text(hjust = 0.5))

print(ft1)
print(ft2)

ggsave(here("img_new/expression_GOIs_neurons.pdf"), plot = ft1, width = 10, height = 6, dpi = 300)
ggsave(here("img_new/Co-expression_neurons_tlr3.pdf"), plot = ft2, width = 20, height = 12, dpi = 600)
```

## Genes of Interest

Set your genes of interest.

```{r genes, echo=FALSE}

genes <- c("Gfral","Tmtc2", "Slc6a15", "Man1a", "Akap13", "Etv1", "Tlr3") # CHANGE

# If you have more than a few genes, then put them in a file "GOIs.txt" (one gene per row) and use the below code
#genes <- readLines(here("GOIs.txt"))
```

## Co-expression GOIs

Run scLink to look at the co-expression of the GOIs. OBS! This step is very resource demanding, make sure to have at least 80 GB RAM on your system. 

```{r scLink, echo=FALSE}

count <- t(expr_matrix_all)

count.norm <- sclink_norm(count, scale.factor = 1e6, filter.genes = FALSE, gene.names = genes)
gc()

networks <- sclink_net(expr = count.norm, ncores = 1, lda = seq(0.5, 0.1, -0.05))

```

# Results from from scLink
```{r}
# Network correlation
networks$cor[1:length(genes),1:length(genes)]
# Network details
net1 = networks$summary[[1]]
net1
```
```{r}

# Assuming `count` is cells x genes and `genes_in_matrix` has your 2 GOIs
Idents(neuron_seurat) <- "neuron_cell_type"

#genes_of_interest <- c("Gfral", "Tlr3")

genes_of_interest <- c("Gfral","Tmtc2", "Slc6a15", "Man1a", "Akap13", "Etv1", "Tlr3")

cell_types <- c("chat1","chat2","chat3","Glu1","Glu2","Glu3","Glu4","Glu5","Glu6",
                 "Glu7","Glu8","Glu9","Glu10","Glu11","Glu12","Glu13","Glu14","Glu15",
                 "GABA1","GABA2","GABA3","GABA4","GABA5","GABA6","GABA7")

# get_expression_stats <- function(seu, cell_type, genes) {
#   subset_obj <- subset(seu, idents = cell_type)
#   mat <- GetAssayData(subset_obj, assay = "RNA", slot = "counts")
# 
#   genes_present <- intersect(genes, rownames(mat))
#   mat <- mat[genes_present, , drop = FALSE]
# 
#   stats <- data.frame(
#     cell_type = cell_type,
#     gene      = genes_present,
#     mean_expr = Matrix::rowMeans(mat),
#     frac_nonzero = Matrix::rowMeans(mat > 0),
#     nonzero_cells = Matrix::rowSums(mat > 0),
#     total_cells = ncol(mat),
#     row.names = NULL
#   )
#   
#   return(stats)
# }

get_expression_stats <- function(seu, cell_type, genes) {
  # make sure default assay is RNA (important in v5)
  DefaultAssay(seu) <- "RNA"

  subset_obj <- subset(seu, idents = cell_type)

  # v5-style: use LayerData instead of GetAssayData
  mat <- LayerData(subset_obj, assay = "RNA", layer = "counts")
  # alternatively: mat <- subset_obj[["RNA"]]$counts

  genes_present <- intersect(genes, rownames(mat))
  mat <- mat[genes_present, , drop = FALSE]

  stats <- data.frame(
    cell_type     = cell_type,
    gene          = genes_present,
    mean_expr     = Matrix::rowMeans(mat),
    frac_nonzero  = Matrix::rowMeans(mat > 0),
    nonzero_cells = Matrix::rowSums(mat > 0),
    total_cells   = ncol(mat),
    row.names     = NULL
  )
  
  return(stats)
}



# Run for all four cell types
expression_results <- do.call(
  rbind,
  lapply(cell_types, function(ct) get_expression_stats(neuron_seurat, ct, genes_of_interest))
)

expression_results



```
# Pairwise Co-expression values per neuron cell type
```{r}
# Run scLink per cell type

# process_neuron_cell_type <- function(neuron_seurat, cell_type, genes) {
#   # Ensure the active identity class is set to the clustering result
#   Idents(neuron_seurat) <- "neuron_cell_type"  
#   
# # Subset the Seurat object to include the specified neuron cell type
# neuron_seurat_subset <- subset(neuron_seurat, idents = cell_type)
# 
# # Extract data layers
# data_layers <- grep("^data$", Layers(neuron_seurat_subset[["RNA"]]), value = TRUE)
# 
# # Extract expression data from all layers and combine
# expr_list <- lapply(data_layers, function(layer) {
#   layer_data <- neuron_seurat_subset[["RNA"]]@layers[[layer]]
# 
#   # Check if layer data is of type S4 and not NULL
#   if (!is.null(layer_data)) {
#     if (inherits(layer_data, "matrix") || inherits(layer_data, "dgCMatrix")) {
#       mat <- GetAssayData(neuron_seurat_subset, assay = "RNA", layer = layer)
# 
#       cat("Processing layer:", layer, "with dimensions:", dim(mat), "\n")
# 
#       # Check if the matrix has two dimensions (genes x cells)
#       if (length(dim(mat)) == 2) {
#         return(mat)
#       } else {
#         warning(paste("Layer", layer, "does not have the expected two dimensions."))
#         return(NULL)  # Skip this layer if it doesn't meet the criteria
#       }
#     } else {
#       warning(paste("Layer", layer, "is not of type S4 or is unexpected. Skipping."))
#       return(NULL)  # Skip this layer if it's not an S4 object
#     }
#   } else {
#     warning(paste("Layer", layer, "is NULL. Skipping."))
#     return(NULL)  # Skip this layer if it's NULL
#   }
# })
# 
# # Remove NULL entries from the list
# expr_list <- Filter(Negate(is.null), expr_list)
# 
# # Check if there are any valid expression matrices
# if (length(expr_list) == 0) {
#   stop(paste("No valid expression matrices found for cell type:", cell_type))
# }
# 
# # Merge all extracted expression matrices (column-wise: cells are in columns)
# expr_matrix <- do.call(cbind, expr_list)
# 
# # Transpose
# count <- t(expr_matrix)
#   
#   # Normalize counts
# count.norm <- sclink_norm(count, scale.factor = 1e6, filter.genes = FALSE, gene.names = genes)
#   
#   gc()
#   
#   # Create networks
#   networks <- tryCatch({
#     sclink_net(expr = count.norm, ncores = 1, lda = seq(0.5, 0.1, -0.05))
#   }, error = function(e) {
#     warning(paste("Error in creating networks for cell type:", cell_type, "\n", e$message))
#     return(NULL)
#   })
#   
#   if (!is.null(networks)) {
#     return(networks$cor[1:length(genes), 1:length(genes)])
#   } else {
#     return(NULL)  # Return NULL if network creation failed
#   }
# }


# process_neuron_cell_type_counts <- function(neuron_seurat, cell_type, genes,
#                                             debug = FALSE) {
#   Idents(neuron_seurat) <- "neuron_cell_type"
#   DefaultAssay(neuron_seurat) <- "RNA"
# 
#   seu <- subset(neuron_seurat, idents = cell_type)
# 
#   # genes x cells, with dimnames (v5-safe)
#   mat <- LayerData(seu, assay = "RNA", layer = "counts")
# 
#   genes_present <- intersect(genes, rownames(mat))
#   if (length(genes_present) < 3) {
#     warning(paste("Cell type", cell_type, ": <3 genes present in counts layer"))
#     return(NULL)
#   }
# 
#   # subset to GOIs and transpose to cells x genes
#   mat_goi  <- mat[genes_present, , drop = FALSE]  # genes x cells
#   count_goi <- t(mat_goi)                         # cells x genes
# 
#   # Drop cells with zero total counts across GOIs (prevents NaNs in normalization)
#   cell_sums <- Matrix::rowSums(count_goi)
#   count_goi <- count_goi[cell_sums > 0, , drop = FALSE]
# 
#   # Drop genes that are all-zero in remaining cells
#   gene_sums <- Matrix::colSums(count_goi)
#   count_goi <- count_goi[, gene_sums > 0, drop = FALSE]
# 
#   if (ncol(count_goi) < 3) {
#     warning(paste("Cell type", cell_type, ": <3 expressed genes after filtering"))
#     return(NULL)
#   }
# 
#   if (debug) {
#     cat("\n==", cell_type, "(counts) ==\n")
#     cat("cells used:", nrow(count_goi), " genes used:", ncol(count_goi), "\n")
#     cat("genes:", paste(colnames(count_goi), collapse = ", "), "\n")
#   }
# 
#   # Normalize counts (scLink expects counts here)
#   count.norm <- sclink_norm(
#     count_goi,
#     scale.factor = scale.factor,
#     filter.genes = FALSE,
#     gene.names = colnames(count_goi)
#   )
# 
#   # Safety: if any non-finite values remain, set to 0 (or you can stop instead)
#   if (any(!is.finite(count.norm))) {
#     warning(paste("Non-finite values after sclink_norm for", cell_type, "- setting to 0"))
#     count.norm[!is.finite(count.norm)] <- 0
#   }
# 
#   networks <- tryCatch(
#     sclink_net(expr = count.norm),
#     error = function(e) {
#       warning(paste("Error in networks for", cell_type, ":", e$message))
#       NULL
#     }
#   )
# 
#   if (is.null(networks)) return(NULL)
# 
#   cor_mat <- networks$cor
#   cor_mat[colnames(count_goi), colnames(count_goi), drop = FALSE]
# }
# 
# 
# # Loop over all neuron cell types
# results <- lapply(neuron_types, function(cell_type) {
#   cat("Processing cell type:", cell_type, "\n")
#   process_neuron_cell_type_counts(neuron_seurat, cell_type, genes)
# })
# 
# # Optionally, name the results for easy identification
# names(results) <- neuron_types

```

```{r}

get_gfral_neighbors <- function(seu, cell_type, anchor="Tlr3",
                                n_neighbors=20, min_cells_gene=5) {

  seu_ct <- subset(seu, idents = cell_type)
  counts <- LayerData(seu_ct, assay="RNA", layer="counts")

  if (!(anchor %in% rownames(counts))) return(character())

  # drop very sparse genes
  keep <- Matrix::rowSums(counts > 0) >= min_cells_gene
  counts <- counts[keep, , drop=FALSE]

  # normalize quickly (CPM + log)
  lib <- Matrix::colSums(counts)
  expr <- log1p(t(t(counts) / lib * 1e4))

  gfral_vec <- expr[anchor, ]
  cors <- apply(expr, 1, function(x) cor(x, gfral_vec, method="spearman"))

  cors <- sort(cors, decreasing=TRUE)
  setdiff(names(cors)[1:n_neighbors], anchor)
}

neighbors_glu4 <- get_gfral_neighbors(neuron_seurat, "GABA1", n_neighbors=20)

#genes <- unique(c(gois, neighbors_glu4))
genes <- neighbors_glu4
length(genes)

Idents(neuron_seurat) <- "neuron_cell_type"
cell_types <- levels(Idents(neuron_seurat))

det_list <- list()

for (ct in cell_types) {
  seu_ct <- subset(neuron_seurat, idents = ct)
  counts <- LayerData(seu_ct, assay = "RNA", layer = "counts")

  genes_present <- intersect(genes, rownames(counts))
  if (length(genes_present) == 0) next

  det <- Matrix::rowMeans(counts[genes_present, , drop = FALSE] > 0)

  det_list[[ct]] <- det
}

```

```{r}
genes_sclink <- genes

# focus on relevant cell types
Idents(neuron_seurat) <- "neuron_cell_type"
seu_sub <- subset(neuron_seurat, idents = c("Glu4", "Glu5"))

counts <- LayerData(seu_sub, assay = "RNA", layer = "counts")

genes_present <- intersect(genes_sclink, rownames(counts))

det_rates <- data.frame(
  gene = genes_present,
  detected_cells = Matrix::rowSums(counts[genes_present, ] > 0),
  total_cells = ncol(counts)
)

det_rates$det_frac <- det_rates$detected_cells / det_rates$total_cells

# add Tlr3 for reference
det_tlr3 <- data.frame(
  gene = "Tlr3",
  detected_cells = sum(counts["Tlr3", ] > 0),
  total_cells = ncol(counts)
)
det_tlr3$det_frac <- det_tlr3$detected_cells / det_tlr3$total_cells

det_rates <- rbind(det_rates, det_tlr3)

# sort by detection rate
det_rates <- det_rates[order(det_rates$det_frac), ]

det_rates

```

```{r}
# Your candidate gene set (put 25 here)
gois <- c(
  "Pcdh9",
  "Gm32857",
  "Tlr3",
  "Trpm3",
  "Ncam2",
  "Cntn4",
  "Ctnnd2",
  "Tenm4",
  "Robo2",
  "Kcnab1",
  "Erbb4",
  "Slc38a11",
  "Akap13",
  "Tmtc2",
  "Etv1",
  "Man1a",
  "Hs3st5",
  "Gabra2",
  "Sntb1",
  "Gm17922",
  "Slc6a15",
  "Mpped2",
  "Gfral",
  "Fras1",
  "Nxn", "Gm9732", "4930503E14Rik", "Crhr2", "Nts"
)



Idents(neuron_seurat) <- "neuron_cell_type"
cell_types <- levels(Idents(neuron_seurat))

# scLink params
min_cells <- 100          # minimum cells per cell type
min_cells_gene <- 5      # gene must be detected in >= this many cells
scale.factor <- 1e4
max_cells <- 1000        # set to NA to disable downsampling (1000 is a good start)
ncores <- 4              # if sclink_net supports ncores in your version

run_sclink_one_fixed_genes <- function(seu, cell_type, gois,
                                       min_cells = 100,
                                       min_cells_gene = 5,
                                       scale.factor = 1e4,
                                       max_cells = 1000,
                                       ncores = 4,
                                       return_pcor = TRUE) {

  # subset to one cell type
  seu_ct <- subset(seu, idents = cell_type)

  # optional downsample cells for speed
  if (!is.na(max_cells) && ncol(seu_ct) > max_cells) {
    set.seed(1)
    keep_cells <- sample(colnames(seu_ct), max_cells)
    seu_ct <- subset(seu_ct, cells = keep_cells)
  }

  # counts matrix: genes x cells
  counts <- LayerData(seu_ct, assay = "RNA", layer = "counts")

  # check cell count
  if (ncol(counts) < min_cells) {
    warning(paste(cell_type, ": too few cells:", ncol(counts)))
    return(NULL)
  }

  # keep GOIs present
  genes_present <- intersect(gois, rownames(counts))
  #if (length(genes_present) < 5) {
  #  warning(paste(cell_type, ": too few GOIs present:", length(genes_present)))
  #  return(NULL)
  #}

  # subset to GOIs
  mat <- counts[genes_present, , drop = FALSE]  # genes x cells

  # drop genes too sparse in this cell type
  n_detect <- Matrix::rowSums(mat > 0)
  keep_genes <- names(n_detect)[n_detect >= min_cells_gene]
  mat <- mat[keep_genes, , drop = FALSE]

  if (nrow(mat) < 5) {
    warning(paste(cell_type, ": too few expressed genes after filtering:", nrow(mat)))
    return(NULL)
  }

  # transpose for scLink: cells x genes
  expr <- t(mat)

  # scLink normalization
  expr_norm <- sclink_norm(expr,
                           scale.factor = scale.factor,
                           filter.genes = FALSE,
                           gene.names = colnames(expr))
  expr_norm[!is.finite(expr_norm)] <- 0

  # run scLink
  net <- tryCatch(
    sclink_net(expr = expr_norm, ncores = ncores),
    error = function(e) {
      warning(paste(cell_type, ": sclink_net failed:", e$message))
      NULL
    }
  )
  if (is.null(net)) return(NULL)

  # extract correlation matrices restricted to GOIs kept
  cor_mat <- net$cor
  gois_used <- intersect(gois, colnames(cor_mat))
  cor_goi <- cor_mat[gois_used, gois_used, drop = FALSE]

  out <- list(
    cell_type = cell_type,
    n_cells_used = nrow(expr_norm),
    n_genes_used = ncol(expr_norm),
    genes_used = gois_used,
    cor_goi = cor_goi
  )

  # partial correlations (optional, but nice for "conditional association")
  if (return_pcor && !is.null(net$pcor)) {
    pcor_mat <- net$pcor
    out$pcor_goi <- pcor_mat[gois_used, gois_used, drop = FALSE]
  }

  # IMPORTANT: do NOT store the full net unless you really need it
  # out$net <- net

  out
}

all_results <- lapply(cell_types, function(ct) {
  message("Processing: ", ct)
  run_sclink_one_fixed_genes(neuron_seurat, ct, gois,
                            min_cells = min_cells,
                            min_cells_gene = min_cells_gene,
                            scale.factor = scale.factor,
                            max_cells = max_cells,
                            ncores = ncores,
                            return_pcor = TRUE)
})
names(all_results) <- cell_types
all_results <- Filter(Negate(is.null), all_results)



```

```{r}
library(pheatmap)

# if you already have it as a matrix:
mat <- all_results$Glu4$cor_goi

# ensure it's numeric matrix
mat <- as.matrix(mat)
storage.mode(mat) <- "numeric"

# optional: order genes exactly how you want
genes <- rownames(mat)
mat <- mat[genes, genes]

# diverging palette centered at 0
bk <- seq(-1, 1, length.out = 101)

pheatmap(mat,
         breaks = bk,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         display_numbers = FALSE,   # remove if too busy
         fontsize_number = 8,
         main = "Gene–gene correlations (scLink cor)",
         filename = here("img_new/gene_gene_corr_scLink_25_genes.pdf"))

gfral_vec <- mat["Gfral", , drop = FALSE]   # 1 x genes

pheatmap(gfral_vec,
         cluster_rows = FALSE,
         cluster_cols = TRUE,       # clusters genes by similarity to Gfral
         display_numbers = TRUE,
         number_format = "%.2f",
         main = "Correlation with Gfral")

gfral_vec2 <- mat[, "Gfral", drop = FALSE]  # genes x 1

pheatmap(gfral_vec2,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         display_numbers = TRUE,
         number_format = "%.2f",
         main = "Correlation with Gfral")

pheatmap(mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         display_numbers = FALSE,
         main = "Gene–gene correlations Glu4",
         filename=here("img_new/heatmap_all_gene_gene_corr_glu4.png"))

library(pheatmap)

# Define reduced, ordered gene list
gene_order <- c(
  "Gfral",
  "Etv1",
  "Man1a",
  "Slc6a15",
  "Akap13",
  "Tmtc2",
  "Tlr3"
)

# Keep only genes that actually exist in the matrix
gene_order <- intersect(gene_order, rownames(mat))

# Subset and order matrix
mat_reduced <- mat[gene_order, gene_order]

pheatmap(
  mat_reduced,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = TRUE,
  number_format = "%.2f",
  fontsize_number = 10,
  main = "Gene–gene correlations Glu 5",
  filename=here("img_new/heatmap_GOIs_gene_gene_corr_glu5.png")
)



```
```{r}
library(Matrix)

get_pairwise_cor_7goi <- function(seu, cell_type, gois_7,
                                  assay="RNA", layer="counts",
                                  min_cells=80,
                                  min_cells_gene=3,
                                  method="spearman",
                                  scale.factor=1e4) {

  seu_ct <- subset(seu, idents = cell_type)
  counts <- LayerData(seu_ct, assay = assay, layer = layer)

  # too few cells
  if (ncol(counts) < min_cells) return(NULL)

  genes_present <- intersect(gois_7, rownames(counts))
  if (length(genes_present) < 2) return(NULL)

  mat <- counts[genes_present, , drop = FALSE]  # genes x cells

  # Normalize (logCPM)
  lib <- Matrix::colSums(mat)
  lib[lib == 0] <- 1
  expr <- log1p(t(t(mat) / lib * scale.factor))  # genes x cells

  # Drop genes detected in too few cells (based on RAW counts)
  n_detect <- Matrix::rowSums(mat > 0)
  keep <- names(n_detect)[n_detect >= min_cells_gene]

  # ensure we keep matrix structure
  expr <- expr[keep, , drop = FALSE]

  # must have at least 2 genes to compute a correlation matrix
  if (nrow(expr) < 2) return(NULL)

  # cor expects observations in rows; we want correlations among genes,
  # so use cells as observations: t(expr) is cells x genes
  cor(as.matrix(t(expr)), method = method, use = "pairwise.complete.obs")
}


library(pheatmap)

gois_7 <- c("Gfral","Tmtc2","Slc6a15","Man1a","Akap13","Etv1","Tlr3")
Idents(neuron_seurat) <- "neuron_cell_type"
cell_types <- levels(Idents(neuron_seurat))

# get correlation matrix per cell type (no scLink)
cor_results <- lapply(cell_types, function(ct) {
  get_pairwise_cor_7goi(neuron_seurat, ct, gois_7,
                        min_cells=80, min_cells_gene=3,
                        method="spearman")
})
names(cor_results) <- cell_types
cor_results <- Filter(Negate(is.null), cor_results)

# build Gfral-vs-others heat matrix
gene_pairs <- lapply(setdiff(gois_7, "Gfral"), function(g) c("Gfral", g))
pair_names <- sapply(gene_pairs, function(p) paste(p[1], p[2], sep="-"))
clusters_to_use <- names(cor_results)

genes_matrix <- do.call(rbind, lapply(gene_pairs, function(pair) {
  sapply(clusters_to_use, function(ct) {
    mat <- cor_results[[ct]]
    if (is.null(mat)) return(NA_real_)
    if (!all(pair %in% rownames(mat))) return(NA_real_)
    as.numeric(mat[pair[1], pair[2]])
  })
}))
rownames(genes_matrix) <- pair_names
colnames(genes_matrix) <- clusters_to_use

# plot
max_val <- max(abs(genes_matrix), na.rm=TRUE)
if (!is.finite(max_val) || max_val == 0) max_val <- 0.25
breaks <- seq(-max_val, max_val, length.out=101)

pheatmap(
  genes_matrix,
  cluster_rows=TRUE,
  cluster_cols=TRUE,
  color=colorRampPalette(c("blue","white","red"))(length(breaks)-1),
  breaks=breaks,
  na_col="gray80",
  angle_col=45,
  main="Gfral co-expression across cell types (Spearman on logCPM)"
)


```

```{r}
# Heatmap code that works with your `all_results` object structure:
# all_results[[cell_type]]$cor_goi is the gene×gene correlation matrix

library(pheatmap)

# ---- 1) Define the 7 GOIs you want in the co-expression heatmap ----
gois_7 <- c("Gfral","Tmtc2","Slc6a15","Man1a","Akap13","Etv1","Tlr3")

# ---- 2) Define the gene pairs: Gfral vs each other GOI ----
gene_pairs <- lapply(setdiff(gois_7, "Gfral"), function(g) c("Gfral", g))
pair_names <- sapply(gene_pairs, function(p) paste(p[1], p[2], sep = "-"))

# ---- 3) Choose which cell types (ALL from scLink results) ----
clusters_to_use <- names(all_results)

# Optional: enforce a nice order (e.g., Glu first, then GABA, then others)
# clusters_to_use <- c(sort(grep("^Glu", clusters_to_use, value=TRUE)),
#                      sort(setdiff(clusters_to_use, grep("^Glu", clusters_to_use, value=TRUE))))

# ---- 4) Build matrix: rows = gene pairs, cols = cell types ----
genes_matrix <- do.call(rbind, lapply(gene_pairs, function(pair) {

  vals <- sapply(clusters_to_use, function(ct) {
    res <- all_results[[ct]]
    if (is.null(res)) return(NA_real_)

    mat <- res$cor_goi
    if (is.null(mat)) return(NA_real_)

    # Check genes present in this cell type's filtered scLink set
    if (!all(pair %in% rownames(mat)) || !all(pair %in% colnames(mat))) {
      return(NA_real_)
    }

    as.numeric(mat[pair[1], pair[2]])
  })

  as.numeric(vals)
}))

rownames(genes_matrix) <- pair_names
colnames(genes_matrix) <- clusters_to_use

# ---- 5) Symmetric color scale around 0 ----
max_val <- max(abs(genes_matrix), na.rm = TRUE)
if (!is.finite(max_val) || max_val == 0) max_val <- 0.25
breaks <- seq(-max_val, max_val, length.out = 101)

# ---- 6) Plot heatmap (keeps NAs as gray so you can see missing pairs) ----
pheatmap(
  genes_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1),
  breaks = breaks,
  main = "Gene Pair Co-expression Across Cell Types (scLink cor)",
  na_col = "gray80",
  angle_col = 45,
  cellheight = 20,
  cellwidth = 20,
  fontsize = 10,
  border_color = "grey70"
)

# ---- OPTIONAL: if you prefer to show 0 instead of NA (not usually recommended) ----
# genes_matrix2 <- genes_matrix
# genes_matrix2[is.na(genes_matrix2)] <- 0
# pheatmap(genes_matrix2, ...same args...)

```

```{r}
# Co-expression heatmap for 7 GOIs across ALL cell types
# Assumes `results` is a named list where each element is a gene×gene correlation matrix
# (e.g., from scLink or any method), with row/colnames = gene symbols.

library(pheatmap)

## 0) Define GOIs (7 genes)
gois <- c("Gfral", "Tmtc2", "Slc6a15", "Man1a", "Akap13", "Etv1", "Tlr3")

## 1) Define gene pairs: Gfral vs each other GOI
gene_pairs <- lapply(setdiff(gois, "Gfral"), function(g) c("Gfral", g))
pair_names <- sapply(gene_pairs, function(p) paste(p[1], p[2], sep = "-"))

## 2) Choose clusters/cell types to include (ALL)
clusters_to_use <- names(all_results)

## 3) Build matrix: rows = gene pairs, cols = clusters
genes_matrix <- do.call(rbind, lapply(seq_along(gene_pairs), function(i) {
  pair <- gene_pairs[[i]]

  vals <- sapply(results[clusters_to_use], function(mat) {
    # mat: correlation matrix for one cell type (or NULL)
    if (is.null(mat)) return(NA_real_)

    # if your `results` stores lists (e.g. list(cor_goi=...)),
    # uncomment and adjust one of these lines:
    # mat <- mat$cor_goi
    # mat <- mat$cor

    # Safety: check genes exist
    if (!all(pair %in% rownames(mat)) || !all(pair %in% colnames(mat))) return(NA_real_)

    as.numeric(mat[pair[1], pair[2]])
  })

  as.numeric(vals)
}))

rownames(genes_matrix) <- pair_names
colnames(genes_matrix) <- clusters_to_use

## 4) Symmetric color scale around 0
max_val <- max(abs(genes_matrix), na.rm = TRUE)
if (!is.finite(max_val) || max_val == 0) max_val <- 0.25  # reasonable fallback
breaks <- seq(-max_val, max_val, length.out = 101)

## 5) Plot heatmap
# Keep NAs as NA so `na_col` works (don’t convert to 0 unless you really want that)
pheatmap(
  genes_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1),
  breaks = breaks,
  main = "Gene Pair Co-expression Across Cell Types",
  na_col = "gray80",
  angle_col = 45,
  cellheight = 20,
  cellwidth = 20,
  fontsize = 10,
  border_color = "grey70"
)

```


```{r}
library(scLink)


count = readRDS(system.file("extdata", "example.rds", package = "scLink"))
genes = readRDS(system.file("extdata", "genes.rds", package = "scLink"))

count.norm = sclink_norm(mat, scale.factor = 1e6, filter.genes = TRUE, n = 500)

networks = sclink_net(expr = count.norm, ncores = 1, lda = seq(0.5, 0.1, -0.05))

networks$cor


net1 = networks$summary[[1]]

```

# Select subset of the above genes for pairwise comparisons
```{r}
gene1 <- "Gfral" # CHANGE
gene2 <- "Slc6a15" # CHANGE  
```

# Pairwise Co-expression values per cell type
```{r}
# Initialize a data frame to hold co-expression values
co_expression_values <- data.frame(CellType = character(), CoExpression = numeric(), stringsAsFactors = FALSE)

# Loop through each entry in results
for (cell_type in names(results)) {
  # Extract the correlation matrix for the current cell type
  cor_matrix <- results[[cell_type]]
  
  # Check if the matrix contains both genes
  if (gene1 %in% rownames(cor_matrix) && gene2 %in% colnames(cor_matrix)) {
    # Extract the co-expression value for the GOIs
    co_expression_value <- cor_matrix[gene1, gene2]
    
    # Append the result to the data frame
    co_expression_values <- rbind(co_expression_values, data.frame(CellType = cell_type, CoExpression = co_expression_value))
  }
}
co_expression_values
```


# Scatter plot
```{r, dev = svg}

# Define clusters of interest
clusters_of_interest <- c("Glu4", "Glu5") # CHANGE

# Define genes of interest
genes_of_interest <- genes

# Define subset
subset <- c(gene1, gene2)

# Extract logcounts for only selected genes
logcounts_data <- as.data.frame(t(as.matrix(logcounts(sce)[genes_of_interest, ])))

# Get metadata with cluster labels
metadata <- colData(sce)[, "predicted_cell_type", drop = FALSE]

# Combine expression data with metadata
subset_data <- cbind(logcounts_data, metadata)

# Filter for the clusters of interest
subset_data <- subset_data[subset_data$predicted_cell_type %in% clusters_of_interest, ]

# Convert to long format for ggplot
subset_data_long <- pivot_longer(subset_data, cols = subset, names_to = "Gene", values_to = "Expression")

# Create scatter plot
g <- ggplot(subset_data, aes(x = gene1, y = gene2, color = predicted_cell_type)) + # CHANGE
  geom_point(alpha = 0.5) +
  facet_wrap(~predicted_cell_type) +
  theme_minimal() +
  labs(title = "Gfral vs. Slc6a15  Expression",  #CHANGE
       x = "Gfral Expression (logcounts)",  #CHANGE
       y = "Slc6a15 Expression (logcounts)") #CHANGE
g

ggsave(here("img/GOIs_scatter_plots_no_exp.svg"), g, bg = 'white')

gene1 <- "Gfral"
gene2 <- "Tlr3"

clusters_of_interest <- c("Glu4", "Glu5")

# Pull only the two genes you want (genes x cells) then transpose to cells x genes
expr2 <- t(as.matrix(logcounts(sce)[c(gene1, gene2), ]))
expr2 <- as.data.frame(expr2)

# Add metadata
expr2$predicted_cell_type <- colData(sce)$predicted_cell_type

# Filter clusters
expr2 <- expr2[expr2$predicted_cell_type %in% clusters_of_interest, , drop = FALSE]

# Plot
g <- ggplot(expr2, aes(x = .data[[gene1]], y = .data[[gene2]], color = predicted_cell_type)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~predicted_cell_type) +
  theme_minimal() +
  labs(
    title = paste(gene1, "vs.", gene2, "Expression"),
    x = paste(gene1, "Expression (logcounts)"),
    y = paste(gene2, "Expression (logcounts)")
  )

g
ggsave(here("img_new/GOIs_scatter_plots_etv1.pdf"), g, bg = "white")

```

# Plot co-expression values for gene1/gene2 per cell type 
```{r message=FALSE, warning=FALSE, dev = svg}
# Extract gene2 correlation values from the gene1 row
# genes_values <- sapply(results, function(x) {
#   val <- x[gene1, gene2]
#   if (is.null(val)) NA else as.numeric(val)  # Convert NULLs to NA and ensure numeric
# })
# 
# # Convert to a 1-row numeric matrix
# genes_matrix <- matrix(genes_values, nrow = 1)
# rownames(genes_matrix) <- "Gfral-Tl3r" #CHANGE
# colnames(genes_matrix) <- names(results)

genes_values <- sapply(results, function(mat) {
  if (is.null(mat)) return(NA_real_)
  if (!(gene1 %in% rownames(mat) && gene2 %in% colnames(mat))) return(NA_real_)
  as.numeric(mat[gene1, gene2])
})

genes_matrix <- matrix(genes_values, nrow = 1)
rownames(genes_matrix) <- paste(gene1, gene2, sep = "-")
colnames(genes_matrix) <- names(results)

# Replace NA values with 0 or use na.color argument in pheatmap
genes_matrix[is.na(genes_matrix)] <- 0  
library(pheatmap)

p <- pheatmap(
  genes_matrix,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  main = paste(gene1, "-", gene2, "Co-expression Across Cell Types"),
  na_col = "gray",
  angle_col = 45,
  cellheight = 40
)

p

ggsave(here("img/heatmap_cluster_coexp_tlr3.svg"), p)
```

# Print co-expression values for only one cluster
```{r message=FALSE, warning=FALSE}
cluster <- "Glu8" #CHANGE
p_cluster <- pheatmap(results[[cluster]])
print(p_cluster)
ggsave(here("img/heatmap_cluster.png"),p_cluster)

```
# Plot co-expression values for gene1/gene2 per cell type 
```{r message=FALSE, warning=FALSE}

library(pheatmap)

## 1. Define gene pairs
gene_pairs <- list(
  c("Gfral", "Tmtc2"),
  c("Gfral", "Tlr3"),
  c("Gfral", "Slc6a15"), 
  c("Gfral", "Man1a"),
  c("Gfral", "Akap13"),
  c("Gfral", "Etv1")
)

pair_names <- sapply(gene_pairs, function(p) paste(p[1], p[2], sep = "-"))

## 2. Choose which clusters to include
##    If you really want ALL clusters, just use:
clusters_to_use <- names(all_results)

## If you only wanted Glu4 & Glu5, you could do:
# clusters_to_use <- intersect(c("Glu4", "Glu5"), names(results))

## 3. Build matrix: rows = gene pairs, cols = clusters
genes_matrix <- do.call(rbind, lapply(gene_pairs, function(pair) {
  vals <- sapply(all_results[clusters_to_use]$cor_goi, function(mat) {
    # mat is one element of `results` (should be a correlation matrix or NULL)

    # If scLink failed for that cell type:
    if (is.null(mat)) return(NA_real_)

    # Safety: check genes exist as row & col names
    if (!all(pair %in% rownames(mat)) || !all(pair %in% colnames(mat))) {
      return(NA_real_)
    }

    # Extract correlation
    as.numeric(mat[pair[1], pair[2]])
  })

  # ensure numeric vector
  as.numeric(vals)
}))

rownames(genes_matrix) <- pair_names
colnames(genes_matrix) <- clusters_to_use

## 4. (Optional) keep NA as NA so pheatmap can color them with na_col
## If you *really* want to force them to 0, you can still do:
# genes_matrix[is.na(genes_matrix)] <- 0

## 5. Symmetric color scale
max_val <- max(abs(genes_matrix), na.rm = TRUE)
if (!is.finite(max_val) || max_val == 0) max_val <- 1  # fallback just in case
breaks <- seq(-max_val, max_val, length.out = 100)

## 6. Dynamic size
n_rows <- nrow(genes_matrix)
n_cols <- ncol(genes_matrix)

genes_matrix[is.na(genes_matrix)] <- 0

## 7. Draw heatmap
p <- pheatmap(
  genes_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1),
  breaks = breaks,
  main = "Gene Pair Co-expression Across Cell Types",
  na_col = "gray",
  angle_col = 45,
  cellheight = 20,
  cellwidth = 20,
  fontsize = 10,
)

p

pheatmap(
  genes_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1),
  breaks = breaks,
  main = "Gene Pair Co-expression Across Cell Types",
  fontsize = 10,
  angle_col = 45,
  cellheight = 20,
  cellwidth = 20,
  width = 12,
  height = 4        
)
p


png(
  here("img/heatmap_cluster_coexp_multiple_genes_trimmed.png"),
  width = 1600,
  height = 500,
  res = 150
)

grid.newpage()
pushViewport(viewport(
  x = 0.5, y = 0.5,
  width = 0.98,
  height = 0.65   # <-- trims top & bottom whitespace
))
grid.draw(p)
dev.off()

####

# Define your gene pairs
gene_pairs <- list(
  c("Gfral", "Tmtc2"),
  c("Gfral", "Tlr3"),
  c("Gfral", "Slc6a15"), 
  c("Gfral", "Man1a"),
  c("Gfral", "Akap13"),
  c("Gfral", "Etv1")
)

# Optional: give rownames for gene pairs (for the heatmap rows)
pair_names <- sapply(gene_pairs, function(p) paste(p[1], p[2], sep = "-"))

# Create matrix: rows = gene pairs, cols = cell types (length of results)
genes_matrix <- do.call(rbind, lapply(gene_pairs, function(pair) {
  vals <- sapply(all_results[[cluster]]$cor_goi, function(x) {
    val <- x[pair[1], pair[2]]
    if (is.null(val)) NA else as.numeric(val)
  })
  vals
}))



pair_names <- sapply(gene_pairs, function(p) paste(p[1], p[2], sep = "-"))

# vector of correlations for Glu4
vals <- sapply(gene_pairs, function(pair) {
  if (!all(pair %in% rownames(results)) || !all(pair %in% colnames(results))) {
    return(NA_real_)
  }
  as.numeric(mat[pair[1], pair[2]])
})

# make it a 6x1 matrix so pheatmap is happy
genes_matrix <- matrix(
  vals,
  nrow = length(vals),
  ncol = 1,
  dimnames = list(pair_names, cluster)
)

genes_matrix

# Set row and column names
rownames(genes_matrix) <- pair_names
colnames(genes_matrix) <- names(results)

# Handle NA values (e.g., replace with 0)
genes_matrix[is.na(genes_matrix)] <- 0

# Symmetric color scale limits
max_val <- max(abs(genes_matrix), na.rm = TRUE)
breaks <- seq(-max_val, max_val, length.out = 100)

# Calculate plot dimensions dynamically
n_rows <- nrow(genes_matrix)
n_cols <- ncol(genes_matrix)
plot_width <- 1 + 0.3 * n_cols + 2  # Added more width (2 instead of 1.5) for color bar
plot_height <- 1 + 0.5 * n_rows

# Save heatmap with enough width to fit color bar
pheatmap(
  genes_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1),
  breaks = breaks,
  main = "Gene Pair Co-expression Across Cell Types",
  na_col = "gray",
  angle_col = 45,
  cellheight = 20,
  cellwidth = 20,
  fontsize = 10,
  filename = here("img_new/heatmap_cluster_coexp_multiple_genes_fullrange.png"),
  width = plot_width,
  height = plot_height
)

# Save heatmap with enough width to fit color bar
pheatmap(
  genes_matrix,
  main = "Gene Pair Co-expression Across Cell Types",
  na_col = "gray",
  angle_col = 45,
  cellheight = 20,
  cellwidth = 20,
  fontsize = 10,
  filename = here("img/heatmap_cluster_coexp_multiple_genes_fullrange.png"),
  width = plot_width,
  height = plot_height
)

# # Cap values to improve color contrast
# cap_value <- 0.1  # You can adjust this depending on your data spread
# genes_matrix_capped <- genes_matrix
# genes_matrix_capped[genes_matrix_capped > cap_value] <- cap_value
# genes_matrix_capped[genes_matrix_capped < -cap_value] <- -cap_value
# 
# # Define color scale
# breaks <- seq(-cap_value, cap_value, length.out = 100)
# 
# # Dynamically calculate plot size
# # Adjust these factors as needed for clarity
# n_rows <- nrow(genes_matrix_capped)
# n_cols <- ncol(genes_matrix_capped)
# plot_width <- 1 + 0.3 * n_cols + 1.5  # +1.5 for space for color bar
# plot_height <- 1 + 0.5 * n_rows
# 
# # Create heatmap
# p <- pheatmap(
#   genes_matrix_capped,
#   color = colorRampPalette(c("blue", "white", "red"))(length(breaks) - 1),
#   breaks = breaks,
#   main = "Gene Pair Co-expression Across Cell Types",
#   na_col = "gray",
#   angle_col = 45,
#   cellheight = 20,
#   cellwidth = 20,
#   fontsize = 10,
#   filename = here("img/heatmap_cluster_coexp_multiple_genes.tiff"),
#   width = plot_width,
#   height = plot_height
# )
# 
# p
# 

```

```{r}
# Required packages
library(Seurat)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stats)

# Your gene list
genes_sclink <- c(
  "Gfral","Tmtc2","Slc6a15","Man1a","Akap13","Etv1",
  "Pcdh9","Trpm3","Robo2","Kcnab1","Ncam2","Sntb1",
  "Ctnnd2","Mpped2","Cntn4","Erbb4","Hs3st5","Fras1",
  "Nxn","Hs3st2","Ralyl","Dmd","Meis1","Ptprd",
  "Kcnmb2","Pde1c","Kcnj3","Spock1","Cntn6","Rora",
  "Pbx3","Calb1","Slc2a13","Dcc","Oxr1","Astn2",
  "Pcsk2","Msi2","Srrm4","Nrxn3","Dlc1"
)

# Genes of interest:
goi <- "Gfral"
test_gene <- "Tlr3"   # the gene you said is nearly absent

# Basic presence counts
counts <- LayerData(neuron_seurat, assay="RNA", layer="counts")
stopifnot(goi %in% rownames(counts))
stopifnot(test_gene %in% rownames(counts))

expr_goi <- counts[goi, ] > 0
expr_test <- counts[test_gene, ] > 0

n_cells <- ncol(counts)
n_goi <- sum(expr_goi)
n_test <- sum(expr_test)
n_both <- sum(expr_goi & expr_test)

cat("Cells total:", n_cells, "\n")
cat(sprintf("%s detected cells: %d (%.3f%%)\n", goi, n_goi, 100 * n_goi / n_cells))
cat(sprintf("%s detected cells: %d (%.3f%%)\n", test_gene, n_test, 100 * n_test / n_cells))
cat("Cells expressing BOTH:", n_both, "\n\n")

# Contingency table and Fisher exact test
tab <- matrix(c(
  sum(expr_goi & expr_test),          # both
  sum(expr_goi & (!expr_test)),       # goi only
  sum((!expr_goi) & expr_test),       # test only
  sum((!expr_goi) & (!expr_test))     # neither
), nrow = 2, byrow = TRUE)
rownames(tab) <- c("Gfral_pos","Gfral_neg")
colnames(tab) <- c("Tlr3_pos","Tlr3_neg")
print(tab)
fisher_res <- fisher.test(tab, alternative = "two.sided")
fisher_res

# Hypergeometric test for overlap (if you want)
# P(X >= n_both) where X ~ Hypergeom(N=n_cells, K=n_test, n=n_goi)
phyper(n_both - 1, m = n_test, n = n_cells - n_test, k = n_goi, lower.tail = FALSE)

# Permutation test for co-expression (preserve marginal detection rates)
set.seed(123)
nperm <- 5000
obs <- n_both
perm_vals <- numeric(nperm)
for(i in seq_len(nperm)){
  perm_test <- sample(expr_test)  # shuffle detection labels
  perm_vals[i] <- sum(expr_goi & perm_test)
}
p_perm <- (sum(perm_vals >= obs) + 1) / (nperm + 1)
cat("Permutation p-value (co-expression >= observed):", p_perm, "\n")

# If counts exist, compute Spearman correlation across all cells and on cells where either gene is expressed
spearman_all <- cor(counts[goi, ], counts[test_gene, ], method = "spearman")
# Spearman only on cells where at least one detected (or both)
sel_cells <- which((counts[goi, ] + counts[test_gene, ]) > 0)
if(length(sel_cells) > 2){
  spearman_sel <- cor(counts[goi, sel_cells], counts[test_gene, sel_cells], method = "spearman")
} else {
  spearman_sel <- NA
}
cat("Spearman (all cells):", spearman_all, "\n")
cat("Spearman (cells with >=1 detection):", spearman_sel, "\n")

# Pseudobulk by cluster: average expression per cluster and correlation
if("seurat_clusters" %in% colnames(neuron_seurat@meta.data)){
  meta <- neuron_seurat@meta.data
  clust_means <- sapply(split(seq_len(n_cells), meta$seurat_clusters), function(idx){
    rowMeans(counts[c(goi, test_gene), idx, drop = FALSE])
  })
  # clust_means will be 2 x nclusters
  if(ncol(clust_means) >= 2){
    cor_pseudobulk <- cor(clust_means[1, ], clust_means[2, ], method = "spearman")
    cat("Pseudobulk Spearman across clusters:", cor_pseudobulk, "\n")
  } else cat("Not enough clusters for pseudobulk correlation\n")
} else cat("No seurat_clusters metadata found; skip pseudobulk\n")

# Visualization: UMAP highlighting cells expressing each gene
# (Assumes seu has reducedDims "umap")
if("umap" %in% names(neuron_seurat@reductions)){
  umap_df <- as.data.frame(Embeddings(neuron_seurat, "umap"))
  umap_df$Gfral_expr <- counts[goi, ] > 0
  umap_df$Tlr3_expr <- counts[test_gene, ] > 0
  p1 <- ggplot(umap_df, aes(x = umap_1, y = umap_2)) +
    geom_point(aes(color = Gfral_expr), size = 0.5) + ggtitle("Gfral detection")
  p2 <- ggplot(umap_df, aes(x = umap_1, y = umap_2)) +
    geom_point(aes(color = Tlr3_expr), size = 0.5) + ggtitle("Tlr3 detection")
  print(p1); print(p2)
} else cat("No UMAP reduction named 'umap' found; skip UMAP plots\n")

# Violin/feature plots for both genes
f <- FeaturePlot(neuron_seurat, features = c(goi, test_gene), cols = c("lightgrey","blue"), ncol = 2)

v <- Seurat::VlnPlot(
  neuron_seurat,
  features = c(goi, test_gene),
  ncol = 2,
  pt.size = 0.1
) & theme(
  axis.text.x = element_text(size = 8)
)

pdf(here("img_new/feature_plots.png"), width = 8, height = 8)
print(f)
dev.off()

pdf(here("img_new/umap_plots.png"), width = 8, height = 8)
print(p1)
print(p2)
dev.off()

pdf(here("img_new/violin_plots.png"), width = 8, height = 8)
print(v)
dev.off()


```

```{r}
library(pheatmap)

gene_pairs <- list(
  c("Gfral", "Tmtc2"),
  c("Gfral", "Tlr3"),
  c("Gfral", "Slc6a15"),
  c("Gfral", "Man1a"),
  c("Gfral", "Akap13"),
  c("Gfral", "Etv1")
)
pair_names <- vapply(gene_pairs, function(p) paste(p[1], p[2], sep="-"), character(1))

clusters_to_use <- names(all_results)
cor_mats <- lapply(all_results[clusters_to_use], function(x) x$cor_goi)

get_pair_cor <- function(mat, g1, g2) {
  if (is.null(mat) || !is.matrix(mat)) return(NA_real_)
  rn <- rownames(mat); cn <- colnames(mat)
  if (is.null(rn) || is.null(cn)) return(NA_real_)
  if (g1 %in% rn && g2 %in% cn) return(as.numeric(mat[g1, g2]))
  if (g2 %in% rn && g1 %in% cn) return(as.numeric(mat[g2, g1]))
  NA_real_
}

genes_matrix <- do.call(rbind, lapply(gene_pairs, function(pair) {
  vapply(cor_mats, function(mat) get_pair_cor(mat, pair[1], pair[2]), numeric(1))
}))
rownames(genes_matrix) <- pair_names
colnames(genes_matrix) <- names(cor_mats)

mat_plot <- genes_matrix                      # keeps NA for display
mat_clust <- genes_matrix
mat_clust[is.na(mat_clust)] <- 0              # only for clustering distances

# Row/col clustering objects from the imputed matrix
row_hc <- hclust(dist(mat_clust))
col_hc <- hclust(dist(t(mat_clust)))

# Symmetric color scale from observed values
max_val <- max(abs(mat_plot), na.rm = TRUE)
if (!is.finite(max_val) || max_val == 0) max_val <- 1
breaks <- seq(-max_val, max_val, length.out = 101)

pheatmap(
  mat_plot,
  cluster_rows = row_hc,
  cluster_cols = col_hc,
  color = colorRampPalette(c("blue","white","red"))(length(breaks) - 1),
  breaks = breaks,
  na_col = "gray",
  main = "Gfral pairwise correlation across cell types",
  angle_col = 45,
  filename=here("img_new/heatmap_all_celltypes_NAs.pdf")
)


```

```{r}

library(Seurat)
library(Matrix)

goi <- "Gfral"
low_genes <- c("Tlr3","Gm9732", "4930503E14Rik", "Nts", "Crhr2")

# helper: run overlap tests for one gene, on a Seurat object (already subset if you want)
coexpr_tests <- function(seu, goi, test_gene, assay="RNA", layer="counts",
                         nperm=5000, seed=123) {

  counts <- LayerData(seu, assay = assay, layer = layer)

  if(!(goi %in% rownames(counts))) stop(paste("Missing goi:", goi))
  if(!(test_gene %in% rownames(counts))) stop(paste("Missing test gene:", test_gene))

  expr_goi  <- counts[goi, ] > 0
  expr_test <- counts[test_gene, ] > 0

  n_cells <- ncol(counts)
  n_goi   <- sum(expr_goi)
  n_test  <- sum(expr_test)
  n_both  <- sum(expr_goi & expr_test)

  # expected overlap under independence
  exp_both <- (n_goi * n_test) / n_cells

  # 2x2 table
  tab <- matrix(c(
    sum(expr_goi & expr_test),          # both
    sum(expr_goi & (!expr_test)),       # goi only
    sum((!expr_goi) & expr_test),       # test only
    sum((!expr_goi) & (!expr_test))     # neither
  ), nrow = 2, byrow = TRUE)
  rownames(tab) <- c("Gfral_pos","Gfral_neg")
  colnames(tab) <- c(paste0(test_gene,"_pos"), paste0(test_gene,"_neg"))

  fisher_res <- fisher.test(tab, alternative = "two.sided")

  # Hypergeometric: P(X >= observed)
  p_hyper <- phyper(n_both - 1, m = n_test, n = n_cells - n_test, k = n_goi, lower.tail = FALSE)

  # Permutation test (shuffle test labels, preserve marginal detection)
  set.seed(seed)
  perm_vals <- replicate(nperm, {
    perm_test <- sample(expr_test)
    sum(expr_goi & perm_test)
  })
  p_perm <- (sum(perm_vals >= n_both) + 1) / (nperm + 1)

  data.frame(
    test_gene = test_gene,
    n_cells = n_cells,
    goi_detected = n_goi,
    test_detected = n_test,
    both_detected = n_both,
    expected_both = exp_both,
    det_frac_goi = n_goi / n_cells,
    det_frac_test = n_test / n_cells,
    fisher_p = fisher_res$p.value,
    fisher_or = unname(fisher_res$estimate),
    fisher_ci_low = fisher_res$conf.int[1],
    fisher_ci_high = fisher_res$conf.int[2],
    p_hyper = p_hyper,
    p_perm = p_perm
  )
}

# --------- Run globally (all neurons in neuron_seurat) ----------
res_all <- do.call(rbind, lapply(low_genes, function(g){
  coexpr_tests(neuron_seurat, goi=goi, test_gene=g)
}))
res_all[order(res_all$det_frac_test), ]

```

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)

df <- res_all %>%
  filter(test_gene %in% c("Gm9732", "4930503E14Rik","Tlr3", "Crhr2", "Nts")) %>%
  mutate(
    test_gene = factor(test_gene, levels = c("Gm9732", "4930503E14Rik","Tlr3", "Crhr2", "Nts")),
    enrichment = both_detected / expected_both
  )

# -------- Panel A: Detection rates --------
pA <- ggplot(df, aes(x = test_gene, y = det_frac_test)) +
  geom_col(width = 0.6, fill = "grey30") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  labs(
    x = "",
    y = "Detection fraction",
    title = "Lowly expressed genes"
  ) +
  theme_classic(base_size = 12) + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

# -------- Panel B: Enrichment (observed / expected) --------
pB <- ggplot(df, aes(x = test_gene, y = enrichment)) +
  geom_col(width = 0.6, fill = "grey30") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10() +
  labs(
    x = "",
    y = "Observed / expected overlap\n(log scale)",
    title = "Co-detection with Gfral"
  ) +
  theme_classic(base_size = 12) + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

# -------- Combine --------
pA + pB
pdf(here("img_new/lowly_expressed_genes_1.pdf"), width = 8, height = 8)
print(pA)
dev.off()

pdf(here("img_new/lowly_expressed_genes_2.pdf"), width = 8, height = 8)
print(pB)
dev.off()

``` 