---
title: "snRNA_rerun_genes"
author: "Kristina Benevides"
date: "2025-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r warning=FALSE, message=FALSE}
library(here)
library(readr)
library(scLink)
library(pheatmap)
library(ggplot2)
library(SingleCellExperiment)
library(Matrix)
library(tidyr)
library(Seurat)
```

## Load the data

```{r load_data}
sce <- readRDS(here("data/sce_neuron.rds"))
neuron_seurat <- readRDS(here("data/neuron_seurat_2.rds"))
expr_matrix_all <- readRDS(here("data/expr_matrix_all.rds"))
results <- readRDS(here("data/neuron_res.rds"))
```

## Check expression of two genes in neurons
```{r message=FALSE, warning=FALSE}

genes_subset <- c("Gfral","Adcyap1") # CHANGE

ft1 <- FeaturePlot(neuron_seurat, features = genes_subset, min.cutoff = "q10")
ft2 <- FeaturePlot(neuron_seurat, features = genes_subset, blend = TRUE, pt.size = 0.1) +
  theme(plot.title = element_text(hjust = 0.5))

print(ft1)
print(ft2)

ggsave(here("img/expression_GOIs_neurons.png"), plot = ft1, width = 10, height = 6, dpi = 300)
ggsave(here("img/Co-expression_GOIs_neurons.png"), plot = ft2, width = 20, height = 12, dpi = 600)
```

## Genes of Interest

Set your genes of interest.

```{r genes, echo=FALSE}

genes <- c("Gfral","Adcyap1", "Gpc3", "Glp1r") # CHANGE

# If you have more genes, like 50-500 then put them in a file "GOIs.txt" (one gene per row) and use the below code
#genes <- readLines(here("GOIs.txt"))
```

## Co-expression GOIs

Run scLink to look at the co-expression Gfral/Adcyap1. OBS! This step is very resource demanding, make sure to have at least 80 GB RAM on your system. 

```{r scLink, echo=FALSE}

count <- t(expr_matrix_all)

count.norm <- sclink_norm(count, scale.factor = 1e6, filter.genes = FALSE, gene.names = genes)
gc()

networks <- sclink_net(expr = count.norm, ncores = 1, lda = seq(0.5, 0.1, -0.05))

```

# Results from from scLink
```{r}
# Network correlation
networks$cor[1:length(genes),1:length(genes)]
# Network details
net1 = networks$summary[[1]]
net1
```

# Select gene for pairwise comparisons

```{r}
gene1 <- "Gfral" # CHANGE
gene2 <- "Adcyap1" # CHANGE  
```

# Pairwise Co-expression values per neuron cell type
```{r}

# Initialize a data frame to hold co-expression values
co_expression_values <- data.frame(CellType = character(), CoExpression = numeric(), stringsAsFactors = FALSE)

# Loop through each entry in results
for (cell_type in names(results)) {
  # Extract the correlation matrix for the current cell type
  cor_matrix <- results[[cell_type]]
  
  # Check if the matrix contains both genes
  if (gene1 %in% rownames(cor_matrix) && gene2 %in% colnames(cor_matrix)) {
    # Extract the co-expression value for the GOIs
    co_expression_value <- cor_matrix[gene1, gene2]
    
    # Append the result to the data frame
    co_expression_values <- rbind(co_expression_values, data.frame(CellType = cell_type, CoExpression = co_expression_value))
  }
}

co_expression_values
```


# Scatter plot
```{r}

# Define clusters of interest
clusters_of_interest <- c("Glu4","Glu5", "Glu8", "Glu10") # CHANGE

# Define genes of interest
genes_of_interest <- genes

# Define subset
subset <- c(gene1, gene2)

# Extract logcounts for only selected genes
logcounts_data <- as.data.frame(t(as.matrix(logcounts(sce)[genes_of_interest, ])))

# Get metadata with cluster labels
metadata <- colData(sce)[, "predicted_cell_type", drop = FALSE]

# Combine expression data with metadata
subset_data <- cbind(logcounts_data, metadata)

# Filter for the clusters of interest
subset_data <- subset_data[subset_data$predicted_cell_type %in% clusters_of_interest, ]

# Convert to long format for ggplot
subset_data_long <- pivot_longer(subset_data, cols = subset, names_to = "Gene", values_to = "Expression")

# Create scatter plot
g <- ggplot(subset_data, aes(x = Gfral, y = Adcyap1, color = predicted_cell_type)) + # CHANGE
  geom_point(alpha = 0.5) +
  facet_wrap(~predicted_cell_type) +
  theme_minimal() +
  labs(title = "Gfral vs. Adcyap1 Expression",  #CHANGE
       x = "Gfral Expression (logcounts)",  #CHANGE
       y = "Adcyap1 Expression (logcounts)") #CHANGE
g

ggsave(here("img/GOIs_scatter_plots.png"), g, bg = 'white')
```

# Plot co-expression values for Gfral/Adcyap1 per cell type 
```{r message=FALSE, warning=FALSE}
# Extract Adcyap1 correlation values from the Gfral row
genes_values <- sapply(results, function(x) {
  val <- x[gene1, gene2]
  if (is.null(val)) NA else as.numeric(val)  # Convert NULLs to NA and ensure numeric
})

# Convert to a 1-row numeric matrix
genes_matrix <- matrix(genes_values, nrow = 1)
rownames(genes_matrix) <- "Gfral-Adcyap1" #CHANGE
colnames(genes_matrix) <- names(results)

# Replace NA values with 0 or use na.color argument in pheatmap
genes_matrix[is.na(genes_matrix)] <- 0  

# Create heatmap
p <- pheatmap(
  genes_matrix,
  cluster_rows = FALSE,  # No need to cluster one row
  cluster_cols = TRUE,   # Cluster cell types
  color = colorRampPalette(c("blue", "white", "red"))(100),
  main = "Gfral-Adcyap1 Co-expression Across Cell Types", #CHANGE
  na_col = "gray", 
  angle_col = 45,
  cellheight = 60
  
)

ggsave(here("img/heatmap_cluster_coexp_dbh.png"), p)
```

# Print co-expression values for only one cluster
```{r message=FALSE, warning=FALSE}
cluster <- "Glu8" #CHANGE
p_cluster <- pheatmap(results[[cluster]])
print(p_cluster)
ggsave(here("img/heatmap_cluster.png"),p_cluster)

```
