---
title: "snRNA-seq QC and pre-processing"
author: "Kristina Benevides"
date: "2024-12-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r, warning=FALSE, message=FALSE}
library(Seurat)
library(DoubletFinder)
library(ggplot2)
library(dplyr)
library(here)
library(tidyverse)
library(parallelDist)
library(doParallel)
library(scran)
library(GSEABase)
library(AUCell)
source("/home/xbenkx/Git/P24-203/scripts/compute_sil.R")
```

# Load data
```{r}
data.directory <- "data/count/"
```

# Create Seurat object
```{r}

names = c("181015_21V", "181015_22S", "181217_APn1_6V", "181217_APn1_12W",
          "181217_APn1_19S","181228_AP-rna-gradient_14V", 
          "181228_AP-rna-gradient_20W", "181228_AP-rna-gradient_27W",
          "181228_AP-rna-gradient_31S", "181228_AP-rna-gradient_36S",
          "181228_AP-rna-wash_9V", "181228_AP-rna-wash_16W", "181228_AP-rna-wash_26S",
          "190222_7W", "190222_37W", "190222_34W", "190222_39S", 
          "190222_40S", "190222_23V", "190222_25V") 

suffix = c("/outs/filtered_feature_bc_matrix")

sample <- gsub(".*_", "", names)
metadata = c() 

for (i in 1:length(names)){

  data.path = paste(paste(data.directory, sample[i], sep = ""), suffix, sep = "") 
  data = Read10X(data.path)
  colnames(data) = paste(sample[i], colnames(data), sep = "_")

  metadata = c(metadata, rep(sample[i], ncol(data)))
  
  if (i == 1) {
    AP.data = data
  } 
  else {
    AP.data = cbind(AP.data, data)
  }
}

metadata = data.frame(metadata)
colnames(metadata) = "sample"
rownames(metadata) = colnames(AP.data)

AP <- CreateSeuratObject(counts = AP.data, meta.data = metadata)
rownames(AP@assays$RNA@layers$counts) <- rownames(AP@assays$RNA)

mito.genes <- grep(pattern = "^mt-", x = rownames(AP@assays$RNA), value = TRUE)
percent.mito <- Matrix::colSums(AP@assays$RNA@layers$counts[mito.genes,])/Matrix::colSums(AP@assays$RNA@layers$counts)

ribo.genes <- grep(pattern = "^Rp", x = rownames(AP@assays$RNA), value = TRUE)
percent.ribo <- Matrix::colSums(AP@assays$RNA@layers$counts[ribo.genes,])/Matrix::colSums(AP@assays$RNA@layers$counts)

# Save data to the Seurat object
mito.ribo = as.data.frame(cbind(percent.mito, percent.ribo))
colnames(mito.ribo) = c("percent.mito", "percent.ribo")

AP <- AddMetaData(object = AP, metadata = mito.ribo)

AP <- subset(AP, subset = percent.mito < .05)

AP.list <- SplitObject(object = AP, split.by = "sample")

```

# Data normalization
```{r}
library(future)
library(future.apply)
plan(multisession, workers = 4) # Adjust workers based on CPU cores

AP.list <- future_lapply(AP.list, function(obj) {
  obj <- NormalizeData(object = obj, verbose = FALSE)
  obj <- FindVariableFeatures(object = obj, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
  obj <- ScaleData(obj, vars.to.regress = "nCount_RNA", verbose = FALSE)
  obj <- RunPCA(obj, npcs = 30, verbose = FALSE)
  return(obj)
})
```

# Data normalization
```{r}
for(i in 1:length(AP.list)) {
  AP.list[[i]] <- NormalizeData(object = AP.list[[i]], verbose = F)
  AP.list[[i]] <- FindVariableFeatures(object = AP.list[[i]], 
                                       selection.method = "vst", nfeatures = 2000, 
                                       verbose = F)
  AP.list[[i]] <- ScaleData(AP.list[[i]], vars.to.regress = "nCount_RNA", verbose = F)
  AP.list[[i]] <- RunPCA(AP.list[[i]], npcs = 30, verbose = F)
}  
saveRDS(AP.list, file = "data/AP_list.rds")
```


# Doublet removal
```{r}

#AP.list <- readRDS("data/AP_list.rds")
cl <- makeCluster(8)
registerDoParallel(cl)

clusterEvalQ(cl, {
  library(DoubletFinder)  # Load the library in each worker
})

AP.list.wodoublets <- foreach(i = 1:length(AP.list)) %dopar% {
  tryCatch({
    set.seed(i)

    # Check if AP.list[[i]] is valid
    if (is.null(AP.list[[i]])) stop("AP.list[[i]] is NULL")
    
    # pK identification
    sweep.res.list_AP <- paramSweep(AP.list[[i]], PCs = 1:10, sct = F)
    sweep.stats_AP <- summarizeSweep(sweep.res.list_AP, GT = F)
    bcmvn_AP <- find.pK(sweep.stats_AP)
    pK <- bcmvn_AP$pK[which.max(bcmvn_AP$BCmetric)]
    pK <- as.numeric(as.character(pK))

    # Doublet removal 
    nExp_poi <- round(0.035 * ncol(AP.list[[i]]@assays$RNA))
    AP.list[[i]] <- doubletFinder(AP.list[[i]], PCs = 1:10, pN = 0.2, pK = pK, 
                                     nExp = nExp_poi, reuse.pANN = F, sct = F)
    
    colnames(AP.list[[i]]@meta.data)[7:8] <- c("pANN", "classifications")
    AP.wodoublets <- subset(AP.list[[i]], subset = classifications == "Singlet")
    list(AP.wodoublets)
  }, error = function(e) {
    message(paste("Error in task ", i, ": ", e$message))
    return(NULL)
  })
}

AP.list.wodoublets <- unlist(AP.list.wodoublets)

names(AP.list.wodoublets) <- names(AP.list)

stopCluster(cl)
saveRDS(AP.list.wodoublets, file = "data/AP_list_wo_doublets.rds")
```

# Data integration
```{r, message=FALSE}
for (i in 1:length(AP.list.wodoublets)) {
 AP.list.wodoublets[[i]] <- NormalizeData(object = AP.list.wodoublets[[i]], verbose = F)
 AP.list.wodoublets[[i]] <- FindVariableFeatures(object = AP.list.wodoublets[[i]],
                                                 selection.method = "vst", 
                                                 nfeatures = 2000, verbose = F)
}

# Integration
AP.anchors <- FindIntegrationAnchors(object.list = AP.list.wodoublets, 
                                     dims = 1:30, anchor.features = 2000)
saveRDS(AP.anchors, file = "data/AP_anchors.rds")
AP <- IntegrateData(anchorset = AP.anchors, dims = 1:30)
saveRDS(AP, file="data/AP_integrated.rds")
```

# Visualization
```{r, warning=FALSE}
# Run the standard workflow for visualization and clustering
AP <- ScaleData(object = AP, verbose = F) 
AP <- RunPCA(object = AP, npcs = 30, verbose = F)
AP <- RunUMAP(object = AP, reduction = "pca", 
    dims = 1:10)
```

# Clustering
```{r, warning=FALSE}
AP <- FindNeighbors(object = AP, reduction = "pca")
#AP <- FindClusters(object = AP, resolution = 0.1, verbose = F)
#AP <- FindClusters(object = AP, resolution = 0.2, verbose = F)
AP <- FindClusters(object = AP, resolution = 0.3, verbose = F)

DimPlot(AP, reduction = "umap", label = T, group.by = "integrated_snn_res.0.3") + 
  theme(legend.position = "none") +
  xlab("UMAP 1") + ylab("UMAP 2")
saveRDS(AP, 
     file = "data/AP_Seurat_obj_tmp0.rds")

AP <- readRDS("data/AP_Seurat_obj_tmp0.rds")
```

# Annotate clusters 
```{r}
# Find cluster markers
cluster_markers <- FindAllMarkers(AP, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

# Save the markers to a file for review
write.csv(cluster_markers, "data/AP_cluster_markers.csv")

# Load marker gene lists for AP and NTS regions
# Replace "ap_markers.csv" and "nts_markers.csv" with the paths to your marker gene files
ap_markers <- read.csv("doc/marker_genes_AP.csv", header = TRUE, stringsAsFactors = FALSE)
nts_markers <- read.csv("doc/marker_genes_NTS.csv", header = TRUE, stringsAsFactors = FALSE)

# Ensure the marker genes are present in your dataset
ap_markers <- intersect(unlist(ap_markers$gene), rownames(AP@assays$RNA))
nts_markers <- intersect(unlist(nts_markers$gene), rownames(AP@assays$RNA))

cluster_avg_expression <- AggregateExpression(AP, features = c(ap_markers, nts_markers), group.by = "seurat_clusters")$RNA

ap_expression <- cluster_avg_expression[ap_markers, ]
nts_expression <- cluster_avg_expression[nts_markers, ]

ap_cluster_scores <- colMeans(ap_expression, na.rm = TRUE)
nts_cluster_scores <- colMeans(nts_expression, na.rm = TRUE)

cluster_annotations <- ifelse(ap_cluster_scores > nts_cluster_scores, "AP", "NTS")
names(cluster_annotations) <- 0:18
mapped_annotations <- cluster_annotations[as.character(AP$seurat_clusters)]
names(mapped_annotations) <- rownames(AP@meta.data)  # Set names to cell IDs
AP$Cluster_Annotation <- mapped_annotations[rownames(AP@meta.data)]

# Visualization with annotations
DimPlot(AP, reduction = "umap", group.by = "Cluster_Annotation", label = TRUE) +
  theme(legend.position = "none") +
  xlab("UMAP 1") + ylab("UMAP 2")

FeaturePlot(AP, features = c("Gfral", "Adcyap1"), min.cutoff = "q10")
FeaturePlot(AP, features = c("Gfral", "Adcyap1"), blend = TRUE, pt.size = 1) +
  theme(plot.title = element_text(hjust = 0.5))
gfral_cells <- WhichCells(AP, expression = Gfral > 0)
adcyap1_cells <- WhichCells(AP, expression = Adcyap1 > 0)

# Find cells that express both genes
overlap_cells <- intersect(gfral_cells, adcyap1_cells)

# Compute percentages
total_cells <- length(Cells(AP))
gfral_pct <- length(gfral_cells) / total_cells * 100
adcyap1_pct <- length(adcyap1_cells) / total_cells * 100
overlap_pct <- length(overlap_cells) / total_cells * 100

# Print results
cat("Percentage of cells expressing Gfral:", round(gfral_pct, 2), "%\n")
cat("Percentage of cells expressing Adcyap1:", round(adcyap1_pct, 2), "%\n")
cat("Percentage of cells expressing both genes:", round(overlap_pct, 2), "%\n")

# Get cell type identities
Idents(AP) <- "cell_type"  # Replace with the correct metadata column for clusters

# Function to compute percentages per cell type
get_overlap_by_cluster <- function(seurat_obj, gene1, gene2) {
  cluster_levels <- levels(seurat_obj)
  results <- data.frame(Cluster = character(), Gfral_pct = numeric(), Adcyap1_pct = numeric(), Both_pct = numeric())
  
  for (cluster in cluster_levels) {
    cells_in_cluster <- WhichCells(seurat_obj, idents = cluster)
    
    gfral_cells <- WhichCells(seurat_obj, expression = !!sym(gene1) > 0)
    adcyap1_cells <- WhichCells(seurat_obj, expression = !!sym(gene2) > 0)
    
    overlap_cells <- intersect(gfral_cells, adcyap1_cells)
    
    gfral_pct <- length(intersect(cells_in_cluster, gfral_cells)) / length(cells_in_cluster) * 100
    adcyap1_pct <- length(intersect(cells_in_cluster, adcyap1_cells)) / length(cells_in_cluster) * 100
    overlap_pct <- length(intersect(cells_in_cluster, overlap_cells)) / length(cells_in_cluster) * 100
    
    results <- rbind(results, data.frame(Cluster = cluster, Gfral_pct = gfral_pct, Adcyap1_pct = adcyap1_pct, Both_pct = overlap_pct))
  }
  
  return(results)
}

# Run the function
overlap_results <- get_overlap_by_cluster(AP, "Gfral", "Adcyap1")
print(overlap_results)

```

# Annotate clusters 

```{r}

non_neurons <- list.files(here("data/csv/non_neurons"), pattern = "*.csv", full.names = TRUE)
neurons <- list.files(here("data/csv/neurons"), pattern = "*.csv", full.names = TRUE)

non_neuron_data <- lapply(non_neurons, read.csv)
names(non_neuron_data) <- basename(non_neurons)
cellex_cutoff <- 0.85
expr_proportion_cutoff <- 10

# Loop through the list and filter each data frame
non_neuron_filtered_data <- lapply(non_neuron_data, function(df) {
  subset(df, cellex_score >= cellex_cutoff & expr_proportion >= expr_proportion_cutoff)
})

output_dir <- "data/marker_genes"  # Directory to save the files
dir.create(output_dir, showWarnings = FALSE)

lapply(seq_along(filtered_marker_data), function(i) {
  output_file <- file.path(output_dir, paste0("marker_genes_", names(filtered_marker_data)[i]))
  write_csv(as.data.frame(filtered_marker_data[[i]]$gene), output_file, col_names = FALSE)
})

```

# Non-neuron annotation

```{r}
# Define directory where your marker gene files are stored
marker_dir <- "data/marker_genes"  # Update with your actual path
neuron_marker_dir <- "data/neuron_marker_genes"
# List all marker gene files
marker_files <- list.files(marker_dir, pattern = "marker_genes_.*\\.csv", full.names = TRUE)
neuron_marker_files <- list.files(neuron_marker_dir, pattern = ".*\\.csv", full.names = TRUE)
markers <- c(marker_files,neuron_marker_files)
# Extract cell type names from the file names
cell_types <- gsub("marker_genes_|\\.csv", "", basename(markers))

# Read marker gene lists and assign cell type names
marker_gene_lists <- lapply(markers, function(file) {
  data <- read.csv(file, header = TRUE, stringsAsFactors = FALSE)
  unlist(data)  # Ensure the marker genes are a character vector
})
names(marker_gene_lists) <- cell_types  # Add cell type names to the list

# Ensure the marker genes are present in your dataset
filtered_marker_genes <- lapply(marker_gene_lists, function(genes) {
  intersect(genes, rownames(AP@assays$RNA))
})

# Calculate average expression for each cell type marker list
cluster_avg_expression <- AggregateExpression(AP, 
                                               features = unlist(filtered_marker_genes), 
                                               group.by = "seurat_clusters")$RNA

# Calculate cluster scores for each cell type
cluster_scores <- lapply(filtered_marker_genes, function(genes) {
  colMeans(cluster_avg_expression[genes, ], na.rm = TRUE)
})

# Annotate clusters based on the highest score for each cell type
cluster_annotations <- c("Neuron", "Astrocytes", "Neuron", "Oligodendrocytes", "Tancyte-like cells", "Neuron", "OPCs", "Microglia", "Neuron", "Neuron", "Neuron", "Neuron", "Ependymal cells", "Neuron", "Neuron", "Endothelial cells", "VLMCs", "Neuron", "Neuron")

names(cluster_annotations) <- gsub("^g", "", colnames(cluster_avg_expression))

# Map annotations back to cells based on cluster IDs
cluster_ids <- AP@meta.data$seurat_clusters  # Extract cluster IDs
mapped_annotations <- cluster_annotations[as.character(cluster_ids)]  # Map annotations using cluster IDs

# Set names of mapped annotations to match cell IDs in AP@meta.data
names(mapped_annotations) <- rownames(AP@meta.data)

# Assign the mapped annotations to AP$Cluster_Annotation
AP$Cluster_Annotation <- mapped_annotations

# Visualization with annotations
DimPlot(AP, reduction = "umap", group.by = "Cluster_Annotation", label = TRUE) +
  theme(legend.position = "none") +
  xlab("UMAP 1") + ylab("UMAP 2")


```

# Neuron annotation
```{r}
# Define the directory containing the marker gene files for neural subtypes
neuron_marker_dir <- "data/neuron_marker_genes"  # Update with your actual path

# List all marker gene files for neural subtypes
neuron_marker_files <- list.files(neuron_marker_dir, pattern = "*\\.csv", full.names = TRUE)

# Extract neural subtype names from the file names
neural_subtypes <- gsub(".csv", "", basename(neuron_marker_files))

# Read marker gene lists and assign neural subtype names
neuron_marker_gene_lists <- lapply(neuron_marker_files, function(file) {
  data <- read.csv(file, header = TRUE, stringsAsFactors = FALSE)
  data$gene  # Extract the 'gene' column
})
names(neuron_marker_gene_lists) <- neural_subtypes 


# Filter marker genes to ensure they are present in your dataset
filtered_neuron_marker_genes <- lapply(neuron_marker_gene_lists, function(genes) {
  intersect(genes, rownames(AP@assays$RNA))
})

# Step 1: Subset the dataset to include only Neurons
neuron_cells <- rownames(AP@meta.data)[AP$Cluster_Annotation == "Neuron"]
neuron_data <- subset(AP, cells = neuron_cells)

neuron_data <- FindNeighbors(neuron_data, dims = 1:30)  # Adjust dims based on elbow plot
neuron_data <- FindClusters(neuron_data, resolution = 0.5)  # Adjust resolution to get 25 clusters

# Check how many clusters were created
table(Idents(neuron_data))

saveRDS(neuron_data, "neuron_data_tmp.rds")
seurat_counts <- neuron_data@assays$RNA
seurat_clusters <- Idents(neuron_data)

# Identify all layers that contain "counts"
count_layers <- grep("^counts", Layers(neuron_data[["RNA"]]), value = TRUE)

# Extract all count layers and store them in a list
count_matrices <- lapply(count_layers, function(layer) {
  LayerData(neuron_data, assay = "RNA", layer = layer)
})

all_genes <- Reduce(union, lapply(count_matrices, rownames))
all_cells <- Reduce(union, lapply(count_matrices, colnames))

# Function to pad matrices so all have the same genes/cells
pad_matrix <- function(mat, all_genes, all_cells) {
  missing_genes <- setdiff(all_genes, rownames(mat))
  missing_cells <- setdiff(all_cells, colnames(mat))
  
  # Add missing genes (rows)
  if (length(missing_genes) > 0) {
    empty_genes <- Matrix(0, nrow = length(missing_genes), ncol = ncol(mat), sparse = TRUE)
    rownames(empty_genes) <- missing_genes
    mat <- rbind(mat, empty_genes)
  }
  
  # Add missing cells (columns)
  if (length(missing_cells) > 0) {
    empty_cells <- Matrix(0, nrow = nrow(mat), ncol = length(missing_cells), sparse = TRUE)
    colnames(empty_cells) <- missing_cells
    mat <- cbind(mat, empty_cells)
  }
  
  # Ensure consistent order
  mat <- mat[all_genes, all_cells, drop = FALSE]
  
  return(mat)
}

# Apply padding to all matrices
count_matrices <- lapply(count_matrices, pad_matrix, all_genes, all_cells)

# Stack matrices together instead of summing
seurat_counts <- do.call(cbind, count_matrices)

# Convert to sparse format
seurat_counts <- as(seurat_counts, "dgCMatrix")

# Check final matrix
print(dim(seurat_counts))  # Should include all genes and cells
print(class(seurat_counts))  # Should be "dgCMatrix"

all.sets <- lapply(names(filtered_neuron_marker_genes), function(x) {
    GeneSet(filtered_neuron_marker_genes[[x]], setName=x)        
})
all.sets <- GeneSetCollection(all.sets)

cbind(nGenes(all.sets))

rankings <- AUCell_buildRankings(seurat_counts,
    plotStats=TRUE, verbose=FALSE)
cell.aucs <- AUCell_calcAUC(all.sets, rankings, aucMaxRank = ceiling(0.10 * nrow(rankings)))
results <- t(assay(cell.aucs))
head(results)



# Step 2: Re-cluster the Subsetted Data
# Normalize and scale the data (if needed)
neuron_data <- NormalizeData(neuron_data)
neuron_data <- FindVariableFeatures(neuron_data)
neuron_data <- ScaleData(neuron_data)
neuron_data <- RunPCA(neuron_data, npcs = 30)
neuron_data <- RunUMAP(neuron_data, dims = 1:30)
neuron_data <- FindNeighbors(neuron_data, dims = 1:30)

# Set a higher resolution to get up to 25 clusters
neuron_data <- FindClusters(neuron_data, resolution = 1.5)  # Adjust resolution as needed

# Step 3: Annotate the New Clusters
# Calculate average expression for neural subtype marker genes
neuron_avg_expression <- AverageExpression(neuron_data, 
                                           features = unlist(filtered_neuron_marker_genes))$RNA

# Calculate cluster scores for each neural subtype
neuron_cluster_scores <- lapply(filtered_neuron_marker_genes, function(genes) {
  colMeans(neuron_avg_expression[genes, , drop = FALSE], na.rm = TRUE)
})

# Combine scores into a data frame
scores_df <- do.call(rbind, neuron_cluster_scores)
colnames(scores_df) <- paste0("g", 0:(ncol(scores_df) - 1))

library(ggplot2)
scores_long <- reshape2::melt(scores_df)

ggplot(scores_long, aes(x = value)) + 
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  facet_wrap(~ Var2, scales = "free_x") + 
  labs(title = "Distribution of Scores for Each Cluster", x = "Score", y = "Frequency")

delta_threshold <- 0.001  # Minimum difference required between top 2 scores

# Assign each cell type to a cluster or leave it unassigned
cluster_assignments <- apply(scores_df, 1, function(scores) {
  sorted_scores <- sort(scores, decreasing = TRUE)  # Sort scores in descending order
  if ((sorted_scores[1] - sorted_scores[2]) >= delta_threshold) {
    names(sorted_scores)[1]  # Assign the cluster with the highest score
  } else {
    NA  # Leave unassigned if the score difference is below the threshold
  }
})

# View the cluster assignments
table(cluster_assignments, useNA = "ifany") 

# Create a result data frame
result <- data.frame(Cluster = colnames(scores_df), AssignedCellType = assigned_cell_types)

# View the assignments
print(result)
# Annotate the new clusters based on the highest score
neuron_cluster_annotations <- apply(as.data.frame(neuron_cluster_scores), 1, function(scores) {
  neural_subtypes[which.max(scores)]
})

names(neuron_cluster_annotations) <- gsub("^g", "", colnames(neuron_avg_expression))

# Map annotations back to cells
new_cluster_ids <- neuron_data@meta.data$seurat_clusters  # Extract new cluster IDs
names(new_cluster_ids) <- rownames(neuron_data@meta.data)

new_mapped_annotations <- neuron_cluster_annotations[as.character(new_cluster_ids)]

names(new_mapped_annotations) <- names(new_cluster_ids)
# Assign the mapped annotations to the neuron data
neuron_data$Neural_Subtype <- new_mapped_annotations

# Handle missing values
neuron_data$Neural_Subtype[is.na(neuron_data$Neural_Subtype)] <- "Unknown"

# Step 4: Visualize the Results
DimPlot(neuron_data, reduction = "umap", group.by = "Neural_Subtype", label = TRUE) +
  theme(legend.position = "right") +
  xlab("UMAP 1") + ylab("UMAP 2")


```

# Load and preprocess ARH-ME data (Campbell et al.)
```{r}

# Load count matrix
campbell.counts <- read.table(gzfile("data/campbell/GSE93374_Merged_all_020816_BatchCorrected_LNtransformed_doubletsremoved_Data.txt.gz"))

# Initialize a Seurat object
campbell <- CreateSeuratObject(counts = campbell.counts)

# Add cell type information to Seurat object
campbell.meta <- read.table(gzfile("data/campbell/GSE93374_cell_metadata.txt.gz"), sep = "\t", header = TRUE)

campbell.meta <- campbell.meta[, c("X1.ID", "X7.clust_all", "All.Cell.Clusters")]
colnames(campbell.meta) <- c("cell.id", "cell.type", "annotation")

mapping <- data.frame(
  cell.type = gsub("\\..*", "", campbell.meta$annotation),  # Extract cluster part (e.g., a01, a02)
  annotation = campbell.meta$annotation,  # Full annotation (e.g., Oligodend3)
  stringsAsFactors = FALSE
)

mapping <- mapping[!duplicated(mapping$cell.type), ]
mapping <- mapping[-nrow(mapping), ]
campbell.meta <- data.frame(apply(campbell.meta, 2, as.character), stringsAsFactors = FALSE)
campbell.meta <- left_join(campbell.meta, mapping, by = "cell.type")
campbell.meta$annotation <- campbell.meta$annotation.y
campbell.meta <- campbell.meta %>%
  select(-annotation.x, -annotation.y)
campbell$cell.id <- rownames(campbell@meta.data)
campbell.meta <- left_join(campbell@meta.data, campbell.meta, by="cell.id")
campbell$cell.type <- campbell.meta$annotation

campbell <- subset(campbell, subset = cell.type %in% c("miss", "a19.ParsTuber1", "a20.ParsTuber2"), invert = TRUE)

# Register parallel backend
library(doParallel)

campbell <- readRDS("data/campbell.rds")
# Normalize data
campbell <- NormalizeData(object = campbell, verbose = F)

Idents(object = campbell) = "cell.type"


num_cores <- 4
registerDoParallel(cores = num_cores)
marker.genes <- c()
marker.genes <- foreach(i = unique(campbell$cell.type))  %dopar% {
  
  markers <- FindMarkers(campbell, 
                        ident.1 = i,
                        test.use = "MAST", only.pos = TRUE, 
                        logfc.threshold = 0.25)
  markers <- rownames(markers)[1:100]
  list(markers)
}


marker.genes <- unlist(marker.genes)

campbell <- ScaleData(object = campbell, verbose = F, features = marker.genes) 
campbell <- RunPCA(object = campbell, npcs = 30, verbose = F, features = marker.genes)
```

# Project labels from ARH-ME to AP
```{r}
AP <- readRDS("data/AP_Seurat_obj_tmp0.rds")
anchors <- FindTransferAnchors(reference = campbell, query = AP, 
                               dims = 1:30, project.query = T, verbose = F, 
                               features = marker.genes)
predictions <- TransferData(anchorset = anchors, refdata = campbell$cell.type, 
                            dims = 1:30, verbose = FALSE, k.weight = 4)

AP$predicted.id <- predictions$predicted.id
AP$prediction.score <- predictions$prediction.score.max

DimPlot(AP, reduction = "umap", label = F, group.by = "predicted.id") + 
  xlab("UMAP 1") + ylab("UMAP 2")
```

# Label microglia
```{r}
AP$clusters <- as.numeric(as.character(AP$integrated_snn_res.0.3))

# Define microglia cluster
AP$clusters[which(AP$predicted.id == "a07.PVMMicro" &
                               AP$prediction.score >= 0.5)] <- 20

major.cell.type <- c("Astrocytes", "Neurons", "Neurons", "Neurons", "Oligodendrocytes", 
                     "Neurons", "Neurons", "Neurons","Neurons", "Neurons",
                     "Neurons", "Neurons", "Neurons", "Neurons", "OPCs",
                     "Ependymal cells", "Tanycyte-like cells", "Endothelial cells", 
                     "VLMCs", "Astrocytes", "Microglia")

AP$cell.type <- major.cell.type[(AP$clusters + 1)]

DimPlot(AP, reduction = "umap", label = F, group.by = "cell.type") + 
  xlab("UMAP 1") + ylab("UMAP 2")

```

# Compute silhouette coefficient
```{r}
pc <- AP@reductions$pca@cell.embeddings[,1:30]
distance <- parDist(pc, method="euclidean")
cluster.id <- AP$cell.type
sil <- compute.sil(x = cluster.id, dist = distance)

AP$silhouette <- sil
AP$remove <- "No"
AP$remove[which(AP$silhouette < 0)] <- "Yes"


DimPlot(AP, reduction = "umap", label = T, group.by = "remove") + 
  theme(legend.position = "none") +
  xlab("UMAP 1") + ylab("UMAP 2")
```

# Remove cells with low silhouette coefficient 
```{r}
AP <- subset(AP, subset = remove == "Yes", invert = T)

AP <- ScaleData(object = AP, verbose = F) 
AP <- RunPCA(object = AP, npcs = 30, verbose = F)
AP <- RunUMAP(object = AP, reduction = "pca", 
    dims = 1:10)

DimPlot(AP, reduction = "umap", label = T, group.by = "cell.type") + 
  theme(legend.position = "none") +
  xlab("UMAP 1") + ylab("UMAP 2")
```

```{r}
saveRDS(AP,
        file = "/projects/mludwig/Ludwig-2021/output/Seurat_objects/AP_Seurat_obj_tmp1.rds")
```

